# Data

## Technical description

### Data Source 1: GHE: Life tables

This dataset is collected and published by the World Health Organization. WHO collects time series of death registration data from as early as the 1950s for around 100 countries. This specific table publishes [annual period life tables for years 2000-2019](https://www.who.int/data/gho/data/indicators/indicator-details/GHO/gho-ghe-life-tables-by-who-region-global-health-estimates).

The dataset can be downloaded as .cvs file and excel file. We chose the .cvs format for easier importing. It has 9310 rows and 34 columns. Among all Indicators in column 2, we are only interested in life expectancy and death rate, so we will only keep rows of these two indicators (ex - expectation of life at age x and nMx: age-specific death rate between ages x and x+n) and drop others. Looking at the columns, there are many duplicated items, such as value and the corresponding value code, and some redundant items such as variable type. Hence, we will only keep the following informative columns: Indicator, Location, Period (year), Dim1 (sex), Dim2 (age group), FactValueNumeric for analysis purpose. After the data cleaning, our dataset has 2660 rows and 6 rows.

### Data Source 2: GHE: Leading causes of death

WHO's Global Health Estimates (GHE) provide the latest available data on [causes of death](https://www.who.int/data/gho/data/themes/mortality-and-global-health-estimates/ghe-leading-causes-of-death) globally, by region and country, and by age, sex and income group. They provide key insights on mortality and morbidity trends to support informed decision-making on health policy and resource allocation.

These estimates are produced using data from multiple sources, including national vital registration data, latest estimates from WHO technical programmes, United Nations partners and inter-agency groups, as well as the Global Burden of Disease and other scientific studies. Before publishing, the GHE are reviewed by WHO Member States through consultation with national focal points and WHO country and regional offices.

The update intervals of this database are not strictly uniform. The specific time panels it includes are: 2000, 2010, 2015, and 2019. It can be exported directly in xlsx format from the website. The main feature columns include: Age Group, Sex Group, Time, Region, and Cause Category. This table records the number of deaths attributed to various causes. It consists of 4 (# of time period) $\times$ 6 (# of regions) sheets, each containing a 215-row (# of causes) $\times$ 16-column (# of gender-age subgroups) dataset table.

This dataset is an aggregate table of death counts, meaning that the recorded data is aggregated across different categorical dimensions rather than representing the raw records of individual deaths. This indicates that the only numeric variable is the death count; all others are categorical variables. This implies that it will assist us in grouping data by different categories, facilitating the exploration of a comprehensible and insightful analysis of global health disparities, and allowing for targeted interventions in key areas.

Please note that the divisions of age, time, gender and region, which are categorical variables with the same meanings, are consistent across both data tables. This ensures that we can obtain research outcomes based on a uniform standard.

## **Research plan**

### Data Source 1: GHE: Life tables

How has the death rate and life expectancy changed over time, globally and regionally?

-   For the global overall trend, we will aggregate the data for each year and make a bar chart showing the death rate and life expectancy at birth from 2000 to 2019.

Does death rate and life expectancy differ by sex?

-   We will aggregate the data for each year by sex and make a side-by-side barchart to compare indicators for male and female, exploring whether there is a gap and whether the gap changes over time.

Do more developed regions have lower death rate, higher life expectancy and larger changes than undeveloped regions?

-   We will aggregate the data for each region and make side-by-side bar charts to compare each region.

Do more developed regions have a more left skewed distribution of death rate on age?

-   First we aggregate the data for each region, and make a histogram of death rate on age groups. 

### Data Source 2: GHE: Leading causes of death

Our research on data source 2 involves the following parts:

#### **a. Temporal Analysis**

-   For this part, we will examine temporal trends in causes of death over the past decade and identify regions or demographic groups experiencing significant changes in mortality rates.

-   **Questions to be answered:** Does the composition of causes of death change over time?

-   **Methods to be implemented:** line chart, Cleveland dot plot.

#### b. **Regional, Age and Gender Disparities**

-   Analyze causes of death by continent, age groups and gender groups, emphasizing descriptive statistics to highlight regional variations. We will explore the influence of socio-economic factors on mortality rates and highlight any discernible patterns or disparities in mortality rates.

-   **Questions to be answered:** What causes death globally and regionally?How do the causes of death differ and appear to be similar across regions, and is this associated with development level? Are there any interaction effects between variables such as age and sex affect the causes of death?

-   **Methods to be implemented:** stacked bar chart for different categories, faceted geographic graphs, PCA analysis with biplot, Interaction effect analysis with Mosaic plot.

By employing EDA and simple statistical tests, this research aims to offer a comprehensible and insightful analysis of global health disparities. The findings from this study can inform targeted public health strategies and contribute to ongoing efforts to improve health outcomes on a global scale.

## Missing values analysis

### Data Source 1:

[To be filled.]{.underline}

### Data Source 2: GHE: Leading causes of death

the death causes in our dataset has different levels. For

```{r}
library(readxl)
library(tidyr)
library(dplyr)
library(ggplot2)

df2 <- read_excel("ghecode2019.xlsx", sheet = "Summary",col_names = TRUE)

df_long <- pivot_longer(df2,cols = Global:WPR, names_to = "region", values_to = "death_num")

filtered_data <- df_long %>%
  filter(region != "Global")

filtered_data <- filtered_data %>%
  mutate(if_zero = ifelse(death_num == 0, 'zero', 'non-zero'))

ggplot(filtered_data, aes(x = if_zero)) +
  geom_bar(stat = "count",fill='cornflowerblue') +
  labs(
    title = "Zero and Non-Zero Proportions on Different Levels of Death Cause Category",
    x = "If Zero",
    y = "Count"
  )+
  facet_wrap(~level,scales = "free_y")+
  theme_minimal()
```

```{r}
file_path <- "ghecode2019_1.xlsx"
df_2019 <- data.frame()
for (sheet_name in excel_sheets(file_path)) {
  current_sheet <- read_excel(file_path, sheet = sheet_name)
  df_2019 <- bind_rows(df_2019, current_sheet)
}

df_2019_long <- pivot_longer(df_2019,cols = '0-28 days':'70+ years', names_to = "age_group", values_to = "death_num")

df_2019_long$age_group <- factor(df_2019_long$age_group, levels = c("0-28 days", "1-59 months", "5-14 years", "15-29 years", "30-49 years", "50-59 years", "60-69 years","70+ years"))

ggplot(df_2019_long, aes(x = region, y = death_num, fill = age_group)) +
  geom_bar(stat = "identity",color=alpha('white',-10)) +
  labs(title = "Death Count in Different Age and Regions for 2019",
       x = "Region",
       y = "Death Count",
       fill = "Age Group")+
  theme_minimal()

ggplot(df_2019_long, aes(x = region, y = death_num, fill = age_group)) +
  geom_bar(stat = "identity",position = "dodge",color=alpha('white',-10)) +
  facet_grid(. ~ sex) +
  labs(title = "Death Count in Different Age and Regions for 2019",
       x = "Region",
       y = "Death Count",
       fill = "Age Group")+
  scale_y_log10() +
  theme_minimal()+
  coord_flip()

# check if is na,两种可能，一种真的没有，一种是missing，先assume真的没有
```

\
